---
title: "Network Analysis of the 2004 Madrid Train Bombing"
author: "Jered Sanchez"
date: "February 5, 2019"
output:
  html_document: default
  pdf_document: default
  word_document: default
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r, echo=FALSE, include=FALSE}
# Load packages
library(tidyverse)
library(igraph)
library(ggraph)

# Read the nodes file into the variable nodes
nodes <- read_csv("data/nodes.csv")

# Read the ties file into the variable ties
ties <- read_csv("data/ties.csv")

# Make the network from the data frame ties and print it
g <- graph_from_data_frame(ties, directed = FALSE, vertices = nodes)
#g

# Give the name "Madrid network" to the network 
g$name <- "Madrid network"
#g$name

# Add node attribute id and print the node `id` attribute
V(g)$id <- seq(1, vcount(g))


```


## Madrid Terrorist Relationship Network

The analysis begins with transforming the data from the files provided into a graph object consisting of nodes and ties, in this case nodes represent terrorists in the network and ties the relationship strenghts between them.

Showing the basic attributes of the network below, we see there are 64 nodes and 243 relationships between them:

```{r}
# Explore the set of nodes
V(g)

# Print the number of nodes

vcount(g)

# Explore the set of ties
E(g)

# Print the number of ties
ecount(g)

```

## Initial Data Visualization

Once we have a general idea of the data and structure of the graph, we can begin to visualize the network. This network has a typical core-periphery structure, with a densely knitted center and a sparser periphery around it.



``` {r, echo = FALSE, warning=FALSE}

# Visualize the network with the Kamada-Kawai layout 
ggraph(g, layout = "with_kk") + 
  # Add an edge link geometry mapping transparency to weight 
  geom_edge_link(aes(alpha = weight)) + 
  # Add a node point geometry
  geom_node_point() +
  # Add a node text geometry, mapping label to id and repelling
  geom_node_text(aes(label = id), repel = TRUE)

# The network has a typical core-periphery structure, with a densely knitted center and a sparser periphery around it.

```

Circular network layout


```{r, echo=FALSE, warning=FALSE}
# Visualize the network in a circular layout
ggraph(g, layout = "in_circle") + 
  # Map tie transparency to its weight
  geom_edge_link(aes(alpha = weight)) + 
  geom_node_point()
```


Grid network layout

```{r, echo = FALSE, warning=FALSE}
# Visualize the network in a grid layout
ggraph(g, layout = "on_grid") + 
  # Map tie transparency to its weight
  geom_edge_link(aes(alpha = weight)) + 
  geom_node_point()
```

## Network Centrality Measures



We can look for the most connected terrorists of the train bombing network using degree centrality. The ranking leader, Jamal Zougam, was in fact directly involved in the bombings and was one of the first to be arrested.
```{r, echo=FALSE, warning=FALSE}
nodes_with_centrality <- nodes %>%
  # Add a column containing the degree of each node
  mutate(degree = degree(g)) %>%
  # Arrange rows by descending degree
  arrange(desc(degree))

# See the result
nodes_with_centrality

```

Several other measures of network centrality can be helpful for analysis. In the following table tie strength, betweenness, and closeness are included. 

### Tie Betweenness

Ties with high betweenness may have considerable influence within a network by virtue of their control over information passing between nodes. Removing them will most disrupt communication between nodes.



``` {r, echo = FALSE, warning=FALSE}
nodes_with_centrality <- nodes %>%
  mutate(
    degree = degree(g),
    # Add a column containing the strength of each node
    strength = strength(g)
  ) %>%
  # Arrange rows by descending strength
  arrange(desc(strength))

# Add betweenness and closeness

nodes_with_centrality <- nodes %>%
  mutate(
    degree = degree(g),
    # Add a column containing the strength of each node
    strength = strength(g),
    betweenness = betweenness(g),
    closeness = closeness(g)
  ) %>%
  # Arrange rows by descending strength
  arrange(desc(closeness))

# See the result
nodes_with_centrality
```

## Final Data Visualization

Once we have an idea of how the network behaves, and the key targets within the network, we can search and visualize the data interactively. This feature allows us to single out a target by selecting the node itself or by the dropdown selector box. You can scoll up and down to zoom on the graph and see the node names.

```{r, echo = FALSE, warning=FALSE}

############################
# Interactive vizualizations
############################

# Load visNetwork
library(visNetwork)

# Convert from igraph to visNetwork
data <- toVisNetworkData(g)

visNetwork(nodes = data$nodes, edges = data$edges) %>%
  # Change the layout to be KK
  visIgraphLayout(layout = "layout_with_kk") %>%
  # Change the options to highlight the nearest nodes and ties
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE)
  

```






